---
title: "Network analytics"
output: html_document
---

# Connecting to the database

```{r}

library(dplyr)
library(purrr)
library(RNeo4j)
library(ggplot2)

  ec_graph <- startGraph("http://localhost:7474/db/data", username = "neo4j", password = "c@mpf1re")
  
```

Some simple stats:

Total number of nodes:
`r unlist(RNeo4j::cypherToList(ec_graph, 'MATCH (n) RETURN count(n)'))`

Total number of awards:
`r unlist(RNeo4j::cypherToList(ec_graph, "MATCH (n:award) RETURN count(n)"))`

Average PIs per grant on all grants

```{r}
aa <- RNeo4j::cypherToList(ec_graph, "match q = (p:person)-[r]-(n:award) RETURN n, count(p) AS people")

aa %>% map(function(x)x$people) %>% unlist %>% mean

```

Average PIs per grant on Geosciences Grants

```{r}
aa <- RNeo4j::cypherToList(ec_graph, "match q = (p:person)-[r]-(n:award)-[s]-(o:organization) WHERE o.directorate =~ '.*Geoscience.*' RETURN n, count(p) AS people")

aa %>% map(function(x)x$people) %>% unlist %>% mean

```


Average PIs per grant on EarthCube Grants

```{r}
aa <- RNeo4j::cypherToList(ec_graph, "match (p:person)-[r]-(n:award) WHERE n.abstract =~ '.*EarthCube.*' RETURN n.id, count(p) AS people")

aa %>% map(function(x)x$people) %>% unlist %>% mean

```

All People by Year:
```{r}

  aa <- RNeo4j::cypherToList(ec_graph, "match (p:person)--(n:award) WHERE exists(n.effectiveDate) RETURN n.id, n.effectiveDate AS date, count(p) AS people") %>% 
  bind_rows %>% 
  mutate(date_code = as.Date(date, format = "%d/%m/%Y"))

ggplot(aa, aes(x = date_code, y = people)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth() +
  scale_y_log10()

```

All EarthCube Grants by Year:

```{r}

  aa <- RNeo4j::cypherToList(ec_graph, "match (p:person)--(n:award) WHERE exists(n.effectiveDate) AND n.abstract =~ '.*EarthCube.*' RETURN n.id, n.effectiveDate AS date, count(p) AS people") %>% 
  bind_rows %>% 
  mutate(date_code = as.Date(date, format = "%d/%m/%Y"))

ggplot(aa, aes(x = date_code, y = people)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth() +
  scale_y_log10()

```
