---
title: "Network analytics"
output:
  html_document:
    self_contined: yes
    code_folding: show
    highlight: pygment
    number_sections: yes
    theme: journal
    toc: yes
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    reference_docx: styles/word_template.docx
    toc: yes
    toc_depth: '3'
---

# Connecting to the database

The database is generated using a separate file.  Initially it ran parallel calls to the `Rscript` command focusing on all NSF awards from the year 2000 onwards. The code is documented in the files `build_award_network` and `batch_run`.  After visiting some more documentation about `neo4j` I switched to a new implementation, detailed in `writecsv.R`, and then the `cypher` file `load_csv.cql`.  With these two files I was able to generate a much more robust implementation, even if it did still take a while on my laptop.

```{r, connect_to_db, results='hide', message=FALSE, warning=FALSE, echo = FALSE}

library(dplyr)
library(purrr)
library(RNeo4j)
library(ggplot2)
library(lubridate)

links <- scan('admin.txt', what = "character")

ec_graph <- RNeo4j::startGraph("http://localhost:7474/db/data", 
                               username = links[1], 
                               password = links[2])

if ('sizes.rds' %in% list.files('data/output')) {
  sizes <- readRDS('data/output/sizes.rds')
} else {
  sizes <- list.files('data/input/awards/', full.names = TRUE) %>%
    map_dbl(function(x)file.info(x)$size)
  saveRDS(sizes, 'data/output/sizes.rds')
}

```

The database is `r round(sum(file.info(list.files('/media/simon/DATA1/neo4j/databases/', all.files = TRUE, recursive = TRUE, full.names = TRUE))$size, na.rm=TRUE)/1000000,0)`MB, but the RNeo4j package makes manipulating it relatively fast.  Some of the functions used here are relatively straightforward, much of it relies on the `MATCH`/`WHERE` paradigm, matching based on either a `(node)` or `[relationship]`.  The challenge in constructing the database has been managing the accession of the relatively large quantity of data (the raw NSF award XML files total `r round(sum(sizes)/1e9, 1)`GB), and also building a reasonable graph.  A challenge with graph construction is reducing the total number of edges between nodes, building reasonably meaningful nodes, and, in particular, ensuring that queries return results that specifically map to awards across longer graphs.  Early iterations of the model failed to include award number in some of the edge elements, which meant that query results could return very large graphs unintentionally since edges weren't clearly tied to award numbers.

# Data Model

 <object type="image/svg+xml" data="images/BroaderDiagram.svg">
 Your browser does not support SVG images
 </object>
 
 The data model uses the [XML Schema for NSF awards](https://github.com/macks22/nsf-award-data/blob/master/docs/nsf-xml-schema-details.md) and is centered around the `award` node.  The `award` node includes `AwardID`, `AwardTitle`, `AbstractNarration`, `AwardAmount`, `AARAAmount`, and `AwardInstrument`.  `AwardEffectiveDate` and `AwardExpirationDate` are expressed as relationships with `Year` nodes.  The same relationship occurs with `MinAmdLetterDate` and `MaxAmdLetterDate`.

# Some Simple Stats

To get a sense of what kinds of things we're looking at, we need to understand the key data objects.  There are nodes for awards, people, institutions, and NSF programs and organizations.  Of the total `r length(sizes)`.  We are interested in patterns of awards since 2000, and we find that XXXX awards meet this condition.

## Unique Node Types

Of the five unique node types, `award` has by far the highest number of nodes, followed by `person`.  While these numbers are likely to be representative of the overall pattern in node size, they don't reflect the true numbers since `people` and `award` are likely to contain a number of duplicates.

```{r, count_node_types, results = 'as-is', message=FALSE, warning=FALSE, echo=FALSE}

node_sizes <- RNeo4j::cypherToList(ec_graph,
                     'START n=node(*)
                      RETURN distinct labels(n) as nodes, count(n) as Count') %>% 
  bind_rows() %>% 
  mutate(Nodes = unlist(nodes)) %>% 
  arrange(desc(Count))

knitr::kable(node_sizes)

```

Duplicates may be identified, in part, through string matches (in particular for `award`), however, for other node types, such as `people`, matches are more difficult to determine, since an individual may both move institutions and share a name with another individual.  Approaches at linking individuals across such moves, and with common names have been undertaken before, where proximity is determined not only by exact name matches, but also by matching networks of co-authors.  In this project we are interested in analysis year by year and will disregard potential effects of unique `person` nodes.

## Unique Relationship Types

Average PIs per grant on all grants, querying the database for all people linked to awards through a single link.  I've added in the program constraint, because it lets us also assess differences between programs.  Because an award might have multiple programs we need to output the awards, group by award ID, but also group by program.

```{r, count_link_types, results = 'as-is', message=FALSE, warning=FALSE, echo=FALSE}

edge_size <- RNeo4j::cypherToList(ec_graph, 
                     'START r=relationship(*)
                      RETURN distinct type(r) as edges, count(r) as count') %>% 
  bind_rows() %>% 
  mutate(edges = unlist(edges)) %>% 
  arrange(desc(count))

knitr::kable(edge_size)
```

Here we see that the links for `funded_by`, which link the `program` and the `award` have a much higher value than the number of awards.  In principle we might expect that each award is funded by only a single `organization`, but in practice we find that many awards are funded by multiple organizations.  Similarly, `awarded_to`, which links an `award` to `people` is also high, because it is again possible for a single award to have multiple awardees at a single institution.  We begin to have fewer relationships per `award` nodes when we link `award` and `institution` with the `administered` relationship, and so on.  This is all to say that the numbers of edges do not seem to indicate any large underlying problems with our model.

To ensure that we were able to follow paths accurately, we embedded the award number into each edge relationship using the `award` property.  This means that for some edges (particularly those with few edges, such as `program_reference`) there are likely to be very many `award` elements, wheras, for those linking an `award` to a `person` there are likely to be few `award` elements in the `awarded_to` relationship.

```{r, awards_per_edge, message=FALSE, warning=FALSE, echo=FALSE, cache=TRUE}
# This tells us how many awards use each path.  From this (awards per edge) we can then
# look at distributions of award numbers per relationship type.

awards_per_edge <- ec_graph %>% 
  cypherToList('START r=relationship(*) RETURN type(r) AS type, SIZE(r.AwardID) AS awards') %>% 
  map(function(x)data.frame(type = x$type, awards = ifelse(length(x$award)==0, 0, x$award), stringsAsFactors=FALSE)) %>% 
  bind_rows() %>% 
  filter(!type %in% c('Awarded_to', 'Funded_by', 'Referenced_by', 
                     'Year_Started', 'Year_Expired', 
                     'Year_Amended_Min', 'Year_Amended_Max'))

ggplot(data = awards_per_edge, aes(x = awards)) + 
  geom_histogram() + 
  facet_wrap(~type, scales = 'free') +
  scale_y_log10() +
  xlab("Number of Awards Along Edge") +
  ylab("Number of Edges")

```

These results are generally as we'd expect here.  There is a 1:1 relationship for awards and the `"funded_by"` and `"administered_by"` edges.  The `"employed_by"` edge represents the total number of awards an individual recieves while employed by an institution, and the `program_element` and `program_reference` connects the `"organization"` to the `"program"` node, so these are effectively internal NSF connections that we'd exepct to see many times.  We'd also expect to see higher values for `"program_reference"` since we'd assume a program is more likely to be referenced than to be the actual funder of a proposal.


```{r, results = 'as-is', message=FALSE, warning=FALSE, echo=FALSE}
# This is a big crunch, it returns the award title, the number of 
pi_perGrant <- ec_graph %>% 
    RNeo4j::cypherToList("match q = 
        (p:Person)-[r:Awarded_to]-(n:Award)-[s:Funded_by]-(o:Directorate) 
        WHERE n.AwardID IN s.AwardID
        RETURN DISTINCT n.AwardTitle           AS title,
               n.AwardInstrument               AS instrument,
               count(DISTINCT n.AwardID)       AS ids,
               collect(DISTINCT o.Directorate) AS directorate, 
               count(DISTINCT p)               AS people")

# direct_map <- read.csv('data/directorate_matches.csv')

split_directorate <- function(x) {
  data.frame(ids = x$ids,
             title = x$title,
             instrument = x$instrument %>% unlist %>% tolower,
             directorate = x$directorate %>% unlist %>% tolower,
             people = x$people,
             stringsAsFactors = FALSE)
  # left_join(y = direct_map, by = "directorate") %>% 
}

pi_perGrant_tbl <- pi_perGrant %>% 
  map(split_directorate) %>% 
  bind_rows()

ggplot(pi_perGrant_tbl) + 
  geom_histogram(aes(x = people)) + 
  facet_wrap(~directorate, scales = 'free')

knitr::kable(pi_perGrant_tbl %>% group_by(directorate) %>% summarise(mean = round(mean(people), 2), sd = round(sd(people),2), count = n()))
```

It's interesting to note the differences here, but, additionally, the fact that many of these.

# The extent of duplicate grants
## How many grants are duplicates?

One challenge is that different institutions submit subawards of the same proposal.  We want to fid out, first, how many exact duplicates there are, how many are the same but with different capitalization or punctuation, and then a distribution of approximate matches for individual grants.

```{r, count_duplicate_awards, results = 'as-is'}

all_titles <- RNeo4j::cypherToList(ec_graph, 
                 "MATCH q = (p:Person)-[r]-(n:Award)-[s]-(o:Directorate)
                  RETURN n.AwardTitle AS title, 
                         n.AwardAmount AS amount, 
                         n.AwardInstrument AS instrument, 
                         count(p) AS people") %>% 
  map(as_data_frame) %>% 
  bind_rows

count_table <- all_titles %>% group_by(people) %>% summarise(count = n())

colnames(count_table) <- c("Title Frequency", "Number of Awards")
knitr::kable(count_table)

```

So, we can see that there are definitely duplicate awards.  It's surprising to me (SJG) that there are so many single institution awards, but I've only ever been on multi-institution awards and so my experience may be atypical.

```{r}

count_table <- all_titles %>% group_by(people, instrument) %>% summarise(count = n())

count_table <- count_table %>% group_by(instrument) %>% filter(n() > 2)

ggplot(count_table, aes(x = people, y = count)) + 
  geom_bar(stat = 'identity') + 
  facet_wrap(~instrument, scales = 'free') +
  scale_y_log10()

```

This gives us a better sense, both of how often multi-institution grant occur, but also how that varies across Award Instrument.

## Looking at Duplicates

So, lets start off by examining the grants with the most duplicate names.

```{r}

mass_award <- all_titles %>% 
  mutate(title, low_title = tolower(title)) %>% 
  group_by(low_title, instrument) %>% 
  filter(n() > 5) %>% summarise(count=n()) %>% 
  arrange(desc(count))

knitr::kable(mass_award)

```

# Average PIs per grant on EarthCube Grants

```{r}
ec_awards <- RNeo4j::cypherToList(ec_graph, 
                           "match (p:Person)-[r]-(n:Award) WHERE n.AbstractNarration =~ '.*EarthCube.*' 
                            RETURN n.AwardID, count(p) AS people")

award_summary <- data.frame(mean = ec_awards %>% map(function(x)x$people) %>% unlist %>% mean,
                            sd   = ec_awards %>% map(function(x)x$people) %>% unlist %>% sd)

```



## All People by Year

```{r, people_by_year, message=FALSE, warning=FALSE}

  people_by_year <- RNeo4j::cypherToList(ec_graph, 
                             "match (p:Person)-[r:Awarded_to]-(n:Award)-[ys:Year_Started]-(y:Year) 
                              WHERE n.AwardID IN ys.AwardID AND y.Year > 1950 AND y.Year < 2018
                              RETURN n.AwardID AS award, y.Year AS year, count(p) AS people") %>% 
  bind_rows

ggplot(people_by_year, aes(x = year, y = people)) + 
  geom_jitter(alpha = 0.5) + 
  geom_smooth(method = "gam", formula = y~ s(x)) +
  scale_y_log10()

```

All EarthCube Grants by Year:

```{r, only_ec_byyear}

  aa <- RNeo4j::cypherToList(ec_graph, 
                             "match (p:Person)-[r:Awarded_to]-(n:Award)-[ys:Year_Started]-(y:Year)
                              WHERE n.AwardID IN ys.AwardID 
                                    AND y.Year > 1950 AND y.Year < 2018 
                                    AND n.AbstractNarration =~ '.*EarthCube.*'
                              RETURN n.AwardID AS award, n.AwardTitle as title, 
                                     y.Year AS year, count(p) AS people") %>% 
  bind_rows %>%
  group_by(title) %>% 
  summarise(total_people = sum(people), year = mean(year))

ggplot(aa, aes(x = year, y = total_people)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "gam", formula = y ~ x, se = TRUE) +
  scale_y_log10()

```

So there's a shift, but the values in the NSF data have so many single PI grants it's hard to really make much of it.

# Diversity of Grant Networks

```{r, eval=TRUE}

pgm_per_yr <- RNeo4j::cypherToList(ec_graph, 
                'MATCH (p:Person)<-[ap:Awarded_to]-(a:Award)-[ay:Year_Started]-(y:Year)
                 MATCH (a)-[apr:Funded_by]-(pr:Program)                
                 WHERE a.AwardID IN apr.AwardID AND
                       a.AwardID IN ay.AwardID
                 WITH DISTINCT p as people,
                      y as year,
                      SIZE(COLLECT(DISTINCT pr.Text)) as program,
                      (CASE WHEN a.AbstractNarration =~ ".*EarthCube.*" 
                            THEN 1 ELSE 0
                       END) AS earthcube
                 RETURN people.EmailAddress as EmailAddress, 
                        people.LastName as LastName, 
                        people.FirstName as FirstName, 
                        year.Year as Year, program, earthcube;') %>% 
  bind_rows()

# Then we clear everyone without an at symbol:
clean_per_yr <- pgm_per_yr %>% filter(grepl("@", EmailAddress))

zeroed_year <- clean_per_yr %>% 
  group_by(EmailAddress) %>% 
  mutate(career_year = Year - min(Year), total_years = max(Year) - min(Year)) %>%
  arrange(EmailAddress, career_year) %>% filter(total_years > 0)

# That last filter kills about 80k people.
  
ggplot(zeroed_year) +
  geom_jitter(aes(x = career_year, y = program), alpha = 0.1) +
  xlab("Year since first grant") +
  ylab("Number of programs applied to")


```
```{r}
ggplot(zeroed_year %>% filter(career_year == 0)) +
  geom_jitter(aes(x = program, y = total_years), alpha = 0.1) +
  xlab("Programs Funding First Grant") +
  ylab("Career Longevity")
```